Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-24 cordillera-cycle.tex
Put no-break spaces after 'a' in new text.

      Although these proxy records were all obtained outside the model
      domain, more regional palaeo-temperature reconstructions spanning over
      the last glacial cycle are lacking. For instance, the Mount Logan ice
      core oxygen isotope record covers only the last 30\,000\,\unit{years}
      (30\,\unit{ka}) and has been interpreted as a proxy for source region
      rather than for palaeo-temperature \citep{Fisher.etal.2004,
      Fisher.etal.2008}. Sea-surface temperatures have been reconstructed
      offshore Vancouver Island from alkenone unsaturation indices over the
      last 16\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Kienast.McKay.2001}, and
      from the Mg/Ca ratio in planktonic foraminifera over the period from
      10 to ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014,
      Taylor.etal.2015}, but these records cover only parts of the last
      glacial cycle.

@@

      Basal topography is bilinearly interpolated from the ETOPO1 combined
      topography and bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009} to the model grids. Sea level is lowered as
      a~function of time based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Ice deformation
      follows temperature and water-content dependent creep
      (Sect.~\ref{sec:icedyn}). Basal sliding follows
      a pseudo-plastic law where the yield stress accounts for till
      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
      Ice shelf calving is calculated based on ice thickness and principle
      strain rates (Sect.~\ref{sec:sliding}). Surface mass balance is
      computed using a~positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Default parameter values are given in
      Table~\ref{tab:params}. Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A sensitivity study
      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
      and basal sliding (Sect.~\ref{sec:sliding}) parameters was performed
      (Table~\ref{tab:sens_params}).

@@

      Ice sheet dynamics are typically modelled using a~combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is combined with the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. The computational
      efficiency of this hybrid scheme nevertheless comes at the cost of
      violating stress conservation in the transition zone where
      gravitational stresses have a double contribution to computed SIA and
      SSA velocities, leading to a positive bias of total ice velocity that
      remains to be quantified.

      Ice deformation is governed by the constitutive law for ice
      \citep{Glen.1952, Nye.1953},
%
\begin{align}
&\label{eqn:glenslaw}
&\vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
\end{align}
%
      where $\vec{\dot{\epsilon}}$ is the strain-rate tensor,
      $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the
      equivalent stress defined by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
      \mathrm{tr}(\vec{\tau}^2)$. The ice softness coefficient, $A$, depends
      on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
      through a piece-wise Arrhenius-type law
      \citep[Eqs.~63--65]{Aschwanden.etal.2012},
%
\begin{align}
&\label{eqn:softness}
&A = E\cdot
\begin{cases}
A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}            & \text{if}\ T_{\text{pa}}  <  T_{\mathrm{c}} \,, \\
A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}} & \text{if}\ T_{\text{pa}} \ge T_{\mathrm{c}} \,,
\end{cases}
\end{align}
%
      where $T_{\text{pa}}$ is the pressure-adjusted ice temperature
      calculated using the Clapeyron relation, ${T_{\text{pa}} = T - \beta
      p}$. $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas
      constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
      $Q_{\mathrm{w}}$, are constant parameters corresponding to values
      measured below and above a critical temperature threshold
      $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
      \citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water
      fraction, $\omega$, is capped at a maximum value of 0.01, above which
      no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
      Greve.1997}. Finally, $E$ is a non-dimensional enhancement factor
      which can take different values, $E_{\text{SIA}}$, in the Shallow Ice
      Approximation (SIA) and $E_{\text{SSA}}$, in the Shallow Shelf
      Approximation (SSA).

@@

      On the other hand, we test the model sensitivity
      (Sect.~\ref{sec:sens}) to different values for the two creep
      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
      enhancement factor, $E_{\text{SIA}}$, as follow.
%
\begin{itemize}
  \item Our \emph{default} configuration used in the control run of the
    sensitivity study and all other simulations in the paper include
    rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
  \item Our \emph{hard ice} configuration include rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to a
    stiffer rheology than that used in the control run.
  \item Our \emph{soft ice} configuration include rheological parameters
    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
    recommended value for ice age polar ice
    \citep[p.~77]{Cuffey.Paterson.2010}.
\end{itemize}
%
      An additional simulation using the ice rheology from
      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} was
      performed, but its results were very similar to that of our default
      run and thus not presented here.

@@

      We also use a constant spatial distribution of the till friction
      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as a
      a~piecewise-linear function of modern bed elevation, with the lowest
      value occuring below modern sea level (0\,\unit{m} above sea level,
      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
      elevation of the highest shorelines
      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
      values span over the range of measured values for glacial till of 18
      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
      for frictional basal conditions associated with discontinuous till
      cover at high elevations, and for a~weakening of till associated with
      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}).

      An additional simulation with a constant till friction angle,
      $\phi=30$\unit{\degree}, corresponding to the average value in
      \citet[p.~268]{Cuffey.Paterson.2010}, was actually performed, but the
      induced variability was small and thus not presented here.

@@

      Using the GRIP ice core palaeo-temperature record as climate forcing
      time-series, the model shows a significant sensitivity to selected ice
      rheological and basal sliding parameters (Fig.~\ref{fig:sens_ts};
      Table~\ref{tab:sens_extrema}) in terms of modelled ice extent, and
      even more so in terms of modelled ice volumes.

      As a direct result of the different scaling factors applied
      (Table~\ref{tab:sens_params}), the resulting simulations show very
      little difference in the modelled glaciated areas corresponding to
      maximum ice volumes during MIS~2, but also during MIS~4
      (Table~\ref{tab:sens_extrema}). However, such can not be said of the
      modelled glaciated area corresponding to minimum ice volume during
      MIS~3. In fact, the extent of the remnant ice cap which persists over
      the Skeena Mountains during this stage shows a significant sensitivity
      to ice rheology of 31\,\unit{\%}, and an even more important
      sensitivity to basal sliding parameters of 62\,\unit{\%}
      (Table~\ref{tab:sens_extrema}).

      The modelled sea-level relevant ice volumes show more variability than
      the modelled glaciated areas (Table~\ref{tab:sens_extrema},
      Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker
      till both result in a thinner ice sheet, while harder ice and stronger
      till result in a thicker ice sheet. For instance, peak ice volume
      during the MIS~2 (LGM) varies by 30\,\unit{\%} between the two
      parametrisations of ice rheology used, and by 21\,\unit{\%} between
      the two parametrisations of basal sliding used. The differences in
      sea-level relevant ice volume are greatest during the MIS~3
      (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
      areal and thickness contributions add up.

@@

      Such deviation from the geological inferences could reflect the fact
      the NARR has difficulties with simulating orographic processes in some
      areas of steep topography \citep{Jarosch.etal.2012}. In a previous
      study \citep{Seguinot.etal.2014}, we have evaluated the performance of
      NARR in forcing constant-climate simulations of the Cordilleran ice
      sheet against that of an observation-based data set
      \citep[WorldClim,][]{Hijmans.etal.2005}. Indeed, the use of NARR in
      these simulations produced slightly different patterns of glaciation
      relative to WorldClim, including more extensive ice cover on the
      Columbia and Rocky mountains \citep[Figs.~6--7]{Seguinot.etal.2014}.
      Our simulations have shown that these differences are mainly caused by
      disparities in the precipitation fields of the two data sets
      \citep[Figs.13--14]{Seguinot.etal.2014}. Over the southern part of our
      model domain, it was shown that the implementation of temperature and
      precipitation downscaling can address these limitations
      \citep[e.g.,][]{Jarosch.etal.2012}. However, extending this
      downscaling method to the entire model domain used in our study could
      be challenging, because the northern part of the model domain is
      characterized by stronger precipitation gradients (Fig.~\ref{fig:atm})
      and includes much fewer weather stations than the south
      \citep{Hijmans.etal.2005}.

      The westerly position of the modelled ice divide and western ice sheet
      margin may also reflect the fact that our model assumes fixed
      modern-day spatial patterns for temperature and precipitation and thus
      does not include feedback mechanisms between ice sheet topography and
      the regional climate. Regarding temperature, it is reasonable to think
      that the cooling was greater inland than near the coast, prohibiting
      melt at the western margin. However, our simulations already produce
      an excess of ice inland. Including such a temperature continentality
      gradient in the model while keeping the precipitation pattern constant
      would thus cause an even greater mismatch between the model results
      and the geologically reconstructed Last Glacial Maximum (LGM) ice
      margins.

@@

      The location of modelled ice-dispersal centres is potentially biased
      by present day ice volumes contained in the ETOPO1 basal topography
      data. Although a reconstruction exist for the thickness of western
      Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013}, the
      most problematic part of the model domain in this respect is by far
      that of the Wrangell and Saint Elias mountains, where ice thicknesses
      up to 1200\,m have been measured by low-frequency radar
      \citep{Rignot.etal.2013}. In this area, located over the USA Canada
      border just north of 60{\degree}N, temperate ice, surge dynamics and
      deep subglacial depressions in the icefield interior pose a
      fundamental challenge to reconstructing basal topography for the
      entire ice cap \citep{Rignot.etal.2013}. Although this causes our
      simulations to overestimate ice surface topography in this region, we
      are not aware of bed topography reconstructions that could be used to
      force the ice sheet model. However, this drawback seem to have
      little effect on modelled Cordilleran ice sheet dynamics. In fact, the
      Wrangell and Saint Elias mountains, heavily glacierized at present,
      host an ice cap for the entire length of both simulations, but that
      ice cap does not appear to be a~major feed to the Cordilleran ice
      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Moreover, with this
      exception of the Wrangell and Saint Elias mountains ice field,
      present-day ice volumes are small relative to the ice volumes
      concerned in our study.

@@

\begin{figure}%
\includegraphics{sens_plot_rheo}
\caption{%
      Ice softness parameter, $A$, as a function of pressure-adjusted
      temperature, $T_{\text{pa}}$, for the default
      \citep{Paterson.Budd.1982}, hard ice \citep[with
      $E_{\text{SIA}}=1$]{Cuffey.Paterson.2010}, and soft ice \citep[with
      $E_{\text{SIA}}=5$]{Cuffey.Paterson.2010} rheologies, using a linear
      scale (top panel) and logarithmic scale (bottom panel). Figure made
      using Eqn.~\ref{eqn:softness} with parameters from
      Table~\ref{tab:sens_params}.}
\label{fig:sens_plot_rheo}%
\end{figure}%

\begin{figure}%
\includegraphics{sens_plot_ntil}
\caption{%
      Effective pressure, $N$, as a function of water content in the till,
      $W$, for the default ($\delta=0.02$, $W_{\text{max}}=2$\,m), hard bed
      ($\delta=0.05$, $W_{\text{max}}=5$\,m), and soft bed ($\delta=0.01$,
      $W_{\text{max}}=1$\,m) sliding parametrisations, using a linear scale
      (top panel) and a logarithmic scale (bottom panel). Calculations are
      made for an ice thickness, $h$, of 1000\,m. Figure made using
      Eqn.~\ref{eqn:ntil} with parameters from Table~\ref{tab:sens_params}
      after \citet[Fig.~1]{Bueler.Pelt.2015}.}
\label{fig:sens_plot_ntil}%
\end{figure}%

@@

\begin{figure*}
\includegraphics{sens_ts}
\caption{%
      Modelled sea-level relevant ice volume through the last 120\,ka in the
      simulation forced by the GRIP paleo-climate record, using default
      parameters (black curves), different ice rheology parameters (top
      panel), and different basal sliding parameters (bottom panel). Gray
      fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
      and MIS~4 according to a global compilation of benthic
      \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
\label{fig:sens_ts}
\end{figure*}

>>>>

      Although these proxy records were all obtained outside the model
      domain, more regional palaeo-temperature reconstructions spanning over
      the last glacial cycle are lacking. For instance, the Mount Logan ice
      core oxygen isotope record covers only the last 30\,000\,\unit{years}
      (30\,\unit{ka}) and has been interpreted as a~proxy for source region
      rather than for palaeo-temperature \citep{Fisher.etal.2004,
      Fisher.etal.2008}. Sea-surface temperatures have been reconstructed
      offshore Vancouver Island from alkenone unsaturation indices over the
      last 16\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Kienast.McKay.2001}, and
      from the Mg/Ca ratio in planktonic foraminifera over the period from
      10 to ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014,
      Taylor.etal.2015}, but these records cover only parts of the last
      glacial cycle.

@@

      Basal topography is bilinearly interpolated from the ETOPO1 combined
      topography and bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009} to the model grids. Sea level is lowered as
      a~function of time based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Ice deformation
      follows temperature and water-content dependent creep
      (Sect.~\ref{sec:icedyn}). Basal sliding follows
      a~pseudo-plastic law where the yield stress accounts for till
      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
      Ice shelf calving is calculated based on ice thickness and principle
      strain rates (Sect.~\ref{sec:sliding}). Surface mass balance is
      computed using a~positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Default parameter values are given in
      Table~\ref{tab:params}. Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A~sensitivity study
      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
      and basal sliding (Sect.~\ref{sec:sliding}) parameters was performed
      (Table~\ref{tab:sens_params}).

@@

      Ice sheet dynamics are typically modelled using a~combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is combined with the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. The computational
      efficiency of this hybrid scheme nevertheless comes at the cost of
      violating stress conservation in the transition zone where
      gravitational stresses have a~double contribution to computed SIA and
      SSA velocities, leading to a~positive bias of total ice velocity that
      remains to be quantified.

      Ice deformation is governed by the constitutive law for ice
      \citep{Glen.1952, Nye.1953},
%
\begin{align}
&\label{eqn:glenslaw}
&\vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
\end{align}
%
      where $\vec{\dot{\epsilon}}$ is the strain-rate tensor,
      $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the
      equivalent stress defined by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
      \mathrm{tr}(\vec{\tau}^2)$. The ice softness coefficient, $A$, depends
      on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
      through a~piece-wise Arrhenius-type law
      \citep[Eqs.~63--65]{Aschwanden.etal.2012},
%
\begin{align}
&\label{eqn:softness}
&A = E\cdot
\begin{cases}
A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}            & \text{if}\ T_{\text{pa}}  <  T_{\mathrm{c}} \,, \\
A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}} & \text{if}\ T_{\text{pa}} \ge T_{\mathrm{c}} \,,
\end{cases}
\end{align}
%
      where $T_{\text{pa}}$ is the pressure-adjusted ice temperature
      calculated using the Clapeyron relation, ${T_{\text{pa}} = T - \beta
      p}$. $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas
      constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
      $Q_{\mathrm{w}}$, are constant parameters corresponding to values
      measured below and above a~critical temperature threshold
      $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
      \citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water
      fraction, $\omega$, is capped at a~maximum value of 0.01, above which
      no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
      Greve.1997}. Finally, $E$ is a~non-dimensional enhancement factor
      which can take different values, $E_{\text{SIA}}$, in the Shallow Ice
      Approximation (SIA) and $E_{\text{SSA}}$, in the Shallow Shelf
      Approximation (SSA).

@@

      On the other hand, we test the model sensitivity
      (Sect.~\ref{sec:sens}) to different values for the two creep
      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
      enhancement factor, $E_{\text{SIA}}$, as follow.
%
\begin{itemize}
  \item Our \emph{default} configuration used in the control run of the
    sensitivity study and all other simulations in the paper include
    rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
  \item Our \emph{hard ice} configuration include rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to
    a~stiffer rheology than that used in the control run.
  \item Our \emph{soft ice} configuration include rheological parameters
    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
    recommended value for ice age polar ice
    \citep[p.~77]{Cuffey.Paterson.2010}.
\end{itemize}
%
      An additional simulation using the ice rheology from
      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} was
      performed, but its results were very similar to that of our default
      run and thus not presented here.

@@

      We also use a~constant spatial distribution of the till friction
      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as
      a~piecewise-linear function of modern bed elevation, with the lowest
      value occuring below modern sea level (0\,\unit{m} above sea level,
      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
      elevation of the highest shorelines
      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
      values span over the range of measured values for glacial till of 18
      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
      for frictional basal conditions associated with discontinuous till
      cover at high elevations, and for a~weakening of till associated with
      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}).

      An additional simulation with a~constant till friction angle,
      $\phi=30$\unit{\degree}, corresponding to the average value in
      \citet[p.~268]{Cuffey.Paterson.2010}, was actually performed, but the
      induced variability was small and thus not presented here.

@@

      Using the GRIP ice core palaeo-temperature record as climate forcing
      time-series, the model shows a~significant sensitivity to selected ice
      rheological and basal sliding parameters (Fig.~\ref{fig:sens_ts};
      Table~\ref{tab:sens_extrema}) in terms of modelled ice extent, and
      even more so in terms of modelled ice volumes.

      As a~direct result of the different scaling factors applied
      (Table~\ref{tab:sens_params}), the resulting simulations show very
      little difference in the modelled glaciated areas corresponding to
      maximum ice volumes during MIS~2, but also during MIS~4
      (Table~\ref{tab:sens_extrema}). However, such can not be said of the
      modelled glaciated area corresponding to minimum ice volume during
      MIS~3. In fact, the extent of the remnant ice cap which persists over
      the Skeena Mountains during this stage shows a~significant sensitivity
      to ice rheology of 31\,\unit{\%}, and an even more important
      sensitivity to basal sliding parameters of 62\,\unit{\%}
      (Table~\ref{tab:sens_extrema}).

      The modelled sea-level relevant ice volumes show more variability than
      the modelled glaciated areas (Table~\ref{tab:sens_extrema},
      Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker
      till both result in a~thinner ice sheet, while harder ice and stronger
      till result in a~thicker ice sheet. For instance, peak ice volume
      during the MIS~2 (LGM) varies by 30\,\unit{\%} between the two
      parametrisations of ice rheology used, and by 21\,\unit{\%} between
      the two parametrisations of basal sliding used. The differences in
      sea-level relevant ice volume are greatest during the MIS~3
      (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
      areal and thickness contributions add up.

@@

      Such deviation from the geological inferences could reflect the fact
      the NARR has difficulties with simulating orographic processes in some
      areas of steep topography \citep{Jarosch.etal.2012}. In a~previous
      study \citep{Seguinot.etal.2014}, we have evaluated the performance of
      NARR in forcing constant-climate simulations of the Cordilleran ice
      sheet against that of an observation-based data set
      \citep[WorldClim,][]{Hijmans.etal.2005}. Indeed, the use of NARR in
      these simulations produced slightly different patterns of glaciation
      relative to WorldClim, including more extensive ice cover on the
      Columbia and Rocky mountains \citep[Figs.~6--7]{Seguinot.etal.2014}.
      Our simulations have shown that these differences are mainly caused by
      disparities in the precipitation fields of the two data sets
      \citep[Figs.13--14]{Seguinot.etal.2014}. Over the southern part of our
      model domain, it was shown that the implementation of temperature and
      precipitation downscaling can address these limitations
      \citep[e.g.,][]{Jarosch.etal.2012}. However, extending this
      downscaling method to the entire model domain used in our study could
      be challenging, because the northern part of the model domain is
      characterized by stronger precipitation gradients (Fig.~\ref{fig:atm})
      and includes much fewer weather stations than the south
      \citep{Hijmans.etal.2005}.

      The westerly position of the modelled ice divide and western ice sheet
      margin may also reflect the fact that our model assumes fixed
      modern-day spatial patterns for temperature and precipitation and thus
      does not include feedback mechanisms between ice sheet topography and
      the regional climate. Regarding temperature, it is reasonable to think
      that the cooling was greater inland than near the coast, prohibiting
      melt at the western margin. However, our simulations already produce
      an excess of ice inland. Including such a~temperature continentality
      gradient in the model while keeping the precipitation pattern constant
      would thus cause an even greater mismatch between the model results
      and the geologically reconstructed Last Glacial Maximum (LGM) ice
      margins.

@@

      The location of modelled ice-dispersal centres is potentially biased
      by present day ice volumes contained in the ETOPO1 basal topography
      data. Although a~reconstruction exist for the thickness of western
      Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013}, the
      most problematic part of the model domain in this respect is by far
      that of the Wrangell and Saint Elias mountains, where ice thicknesses
      up to 1200\,m have been measured by low-frequency radar
      \citep{Rignot.etal.2013}. In this area, located over the USA Canada
      border just north of 60{\degree}N, temperate ice, surge dynamics and
      deep subglacial depressions in the icefield interior pose
      a~fundamental challenge to reconstructing basal topography for the
      entire ice cap \citep{Rignot.etal.2013}. Although this causes our
      simulations to overestimate ice surface topography in this region, we
      are not aware of bed topography reconstructions that could be used to
      force the ice sheet model. However, this drawback seem to have
      little effect on modelled Cordilleran ice sheet dynamics. In fact, the
      Wrangell and Saint Elias mountains, heavily glacierized at present,
      host an ice cap for the entire length of both simulations, but that
      ice cap does not appear to be a~major feed to the Cordilleran ice
      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Moreover, with this
      exception of the Wrangell and Saint Elias mountains ice field,
      present-day ice volumes are small relative to the ice volumes
      concerned in our study.

@@

\begin{figure}%
\includegraphics{sens_plot_rheo}
\caption{%
      Ice softness parameter, $A$, as a~function of pressure-adjusted
      temperature, $T_{\text{pa}}$, for the default
      \citep{Paterson.Budd.1982}, hard ice \citep[with
      $E_{\text{SIA}}=1$]{Cuffey.Paterson.2010}, and soft ice \citep[with
      $E_{\text{SIA}}=5$]{Cuffey.Paterson.2010} rheologies, using a~linear
      scale (top panel) and logarithmic scale (bottom panel). Figure made
      using Eqn.~\ref{eqn:softness} with parameters from
      Table~\ref{tab:sens_params}.}
\label{fig:sens_plot_rheo}%
\end{figure}%

\begin{figure}%
\includegraphics{sens_plot_ntil}
\caption{%
      Effective pressure, $N$, as a~function of water content in the till,
      $W$, for the default ($\delta=0.02$, $W_{\text{max}}=2$\,m), hard bed
      ($\delta=0.05$, $W_{\text{max}}=5$\,m), and soft bed ($\delta=0.01$,
      $W_{\text{max}}=1$\,m) sliding parametrisations, using a~linear scale
      (top panel) and a~logarithmic scale (bottom panel). Calculations are
      made for an ice thickness, $h$, of 1000\,m. Figure made using
      Eqn.~\ref{eqn:ntil} with parameters from Table~\ref{tab:sens_params}
      after \citet[Fig.~1]{Bueler.Pelt.2015}.}
\label{fig:sens_plot_ntil}%
\end{figure}%

@@

\begin{figure*}
\includegraphics{sens_ts}
\caption{%
      Modelled sea-level relevant ice volume through the last 120\,ka in the
      simulation forced by the GRIP paleo-climate record, using default
      parameters (black curves), different ice rheology parameters (top
      panel), and different basal sliding parameters (bottom panel). Gray
      fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
      and MIS~4 according to a~global compilation of benthic
      \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
\label{fig:sens_ts}
\end{figure*}
